Experience Report: Haskell in Computational Biology

### 2012-03-09 mrfy_experience_report.tex
fixed git conflict

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins. 
\remark{``Homology'' means ``share a common ancestor.  I hope we can
avoid this story. ---NR}

@@

\emph{[GIT CONFLICT HERE:
 simply implements the Viterbi algorithm for computing the optimal parse 
of a protein sequence onto a hidden Markov model.
Thus, the graphical model of 
a HMMER hidden Markov model is exactly that shown in Figure~\ref{plan7}.
In 
contrast, MRFy builds on the approach of SMURF~\cite{Menke:2010ti} and 
SMURFLite~\cite{Daniels:2012}, our previous work.
Like SMURF and SMURFLite, MRFy 
pays attention to longer-range structural interactions than a simple hidden 
Markov model can understand.
]}

@@

These non-local interactions cause what had been a simple hidden Markov model 
to become a Markov random field.
SMURF and SMURFLite attempted to compute the 
optimal parse of a protein sequence onto the directed graph representing a 
Markov random field using multidimensional dynamic programming.
SMURF's 
computational complexity quickly exploded with complex protein structures, 
while SMURFLite limited its computational complexity by also limiting the 
complexity of protein structure it would understand.
The algorithm underlying 
MRFy grew out of an understanding of these limitations.

>>>>

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins. 
\remark{``Homology'' means ``share a common ancestor.''  I hope we can
avoid this story. ---NR}

@@

@@

These non-local interactions cause what had been a simple hidden Markov model 
to become a Markov random field.
SMURF~\cite{Menke:2010ti} and SMURFLite~\cite{Daniels:2012}, our previous work,
attempted to compute the 
optimal parse of a protein sequence onto the directed graph representing a 
Markov random field using multidimensional dynamic programming.
SMURF's 
computational complexity quickly exploded with complex protein structures, 
while SMURFLite limited its computational complexity by also limiting the 
complexity of protein structure it would understand.
The algorithm underlying 
MRFy grew out of an understanding of these limitations.
